---
title: "Merging HINTS 5 with R"
author: 
  - name          : "Michael Laviolette, PhD MPH" 
    email         : "statman54@gmail.com"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
```

# Introduction

This report presents results from merging data from HINTS 5, Cycles 1, 2, and 3. For Cycle 3 three survey modalities were used. NCI recommends that analysts assess whether the variables of interest differ across modality groups. If differences are not significant, Cycle 3 can be analyzed as a combined sample; otherwise the analyst should control for groups. Both scenarios are considered using as examples the variables `SeekHealthInfo` and `ChanceAskQuestions`. See the script `HINTS-merge5.R` for details of the merging. Results from SAS are in the Excel files 'no_differences.xlsx` and `with_differences.xlsx`.

```{r load-pkgs, message=FALSE}
library(dplyr)
library(srvyr)
library(kableExtra)
source("HINTS-merge5.R")
percent <- function(x, decimals = 4) round(100 * x, decimals)
```

All results agree with those obtained using SAS. 

## `SeekHealthInfo` assuming no group differences

```{r info-no-diff}
hints5_svy_no_diff %>% 
  group_by(survey, SeekHealthInfo) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE)) %>% 
  mutate_at(vars(starts_with("pct")), percent) %>% 
  kable() %>% 
  kable_styling()

```

## `ChanceAskQuestions` assuming no group differences

```{r questions-no-diff}
hints5_svy_no_diff %>% 
  group_by(survey, ChanceAskQuestions) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE)) %>% 
  mutate_at(vars(starts_with("pct")), percent) %>% 
  kable() %>% 
  kable_styling()

```

## `SeekHealthInfo` controlling for group differences

Note that Cycle 3 is now subgrouped by modality. Since Cycles 1 and 2 have no subgroups, those results are unchanged.

```{r info-with-diff}
hints5_svy_with_diff %>% 
  group_by(survey, treatment, SeekHealthInfo) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE)) %>% 
  mutate_at(vars(starts_with("pct")), percent) %>% 
  kable() %>% 
  kable_styling()

```

## `ChanceAskQuestions` controlling for group differences

```{r questions-with-diff}
hints5_svy_with_diff %>% 
  group_by(survey, treatment, ChanceAskQuestions) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE)) %>% 
  mutate_at(vars(starts_with("pct")), percent) %>% 
  kable() %>% 
  kable_styling()

```

# Conclusion

Merging of multiple HINTS cycles is feasible with R and produces the same results as SAS.

