---
title: "HINTS with R"
author: 
  - name          : "Michael Laviolette, PhD MPH" 
    email         : "statman54@gmail.com"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 4)
```

Page citations refer to ["Overview of the Hints 5 Cycle 3 Survey and Data Analysis Recommendations"](https://hints.cancer.gov/dataset/HINTS5_Cycle3_SAS_03112020.zip).

```{r load-pkgs, message=FALSE}
library(broom)
library(dplyr)
library(haven)
library(purrr)
library(srvyr)
library(survey)
library(tibble)
library(tidyr)
library(kableExtra)
```

## Import and recode data, pp. 14-15

```{r}
edu_lbl <- c("Less than high school",
             "12 years or completed high school",
             "Some college",
             "College graduate or higher")
health_lbl <- c("Excellent", "Very good", "Good", "Fair", "Poor")
group_lbl <-  c("Paper only", "Web option", "Web bonus")
# convenience function to show percentages to four places like SAS
percent <- function(x, decimals = 4) round(100 * x, decimals)
# import SAS data set contained in .zip file
hints5_3 <- read_sas(unz("HINTS5_Cycle3_SAS_03112020.zip",
                         "hints5_cycle3_public.sas7bdat")) %>% 
  mutate(gender = factor(GenderC, 1:2, c("Male", "Female")),
         # recode negative GeneralHealth values to missing
         GeneralHealth = if_else(GeneralHealth < 0, NA_real_, GeneralHealth),
         # collapse education to four levels
         edu = case_when(Education %in% 1:2 ~ 1,
                         Education == 3     ~ 2,
                         Education %in% 4:5 ~ 3,
                         Education %in% 6:7 ~ 4,
                         TRUE ~ NA_real_)) %>% 
  # change edu and Treatment_H5C3 from numeric to factor
  mutate_at("edu", factor, 1:4, edu_lbl) %>% 
  mutate_at("Treatment_H5C3", factor, 1:3, group_lbl) %>% 
  # "No" must be the referent level of SeekCancerInfo to model probability of
  # "Yes". By default the first level of a factor is the referent level, so we
  # reverse the order from the data.
  mutate_at("SeekCancerInfo", factor, 2:1, c("No", "Yes"))
```

## Assessing differences across groups, binary outcome, p. 8

```{r}
hints5_3_grp <- as_survey_rep(hints5_3, weights = "nwgt0", 
                              repweights = paste0("nwgt", 1:150), 
                              type = "JK1", scale = 49/50, mse = TRUE) 
model01 <- svyglm(SeekCancerInfo ~ Treatment_H5C3, hints5_3_grp, 
                  family = "quasibinomial") 
regTermTest(model01, "Treatment_H5C3")

```

## Assessing differences across groups, continuous outcome, p. 8

```{r}
model02 <- svyglm(GeneralHealth ~ Treatment_H5C3, hints5_3_grp)
regTermTest(model02, "Treatment_H5C3")

```

# Replicate weights variance estimation method
## Frequency table and chi-square test, p. 16

```{r, warning = FALSE}
# construct replicate weights survey object
hints5_3_rep <- as_survey_rep(hints5_3, weights = "TG_all_FINWT0",
                              repweights = paste0("TG_all_FINWT", 1:50),
                              type = "JK1", scale = 49/50, mse = TRUE)

```

```{r, warning=FALSE}
tbl_1 <- hints5_3_rep %>% 
  group_by(edu, gender) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE, vartype = c("se", "ci"))) %>% 
  drop_na() %>% 
  mutate_at(vars(starts_with("pct")), percent)

tbl_1 %>% 
  kable() %>% 
  kable_styling()
```

```{r}
chisq_test <- svychisq(~ edu + gender, hints5_3_rep, statistic = "Wald")
chisq_test$statistic
chisq_test$parameter

chisq_test_adj <- svychisq(~ edu + gender, hints5_3_rep, statistic = "adjWald")
chisq_test_adj$statistic
chisq_test_adj$parameter
```

## Logistic regression, pp. 17-18

```{r}
model03 <- svyglm(SeekCancerInfo ~ gender + edu, hints5_3_rep, 
                  family = "quasibinomial")
model03$df.resid
```

### Analysis of effects, p. 17
```{r}
regTermTest(model03, "gender", df = 49)
regTermTest(model03, "edu", df = 49)

```

### Odds ratios, p. 18

```{r}
exp(coef(model03)) 
exp(confint(model03, ddf = 49)) 

```

## Linear regression, pp. 18-19
### Estimates and model effects tests, p.19

```{r}
model04 <- svyglm(GeneralHealth ~ gender + edu, hints5_3_rep)
tidy(model04) %>% 
  kable() %>% 
  kable_styling()
```

```{r}
regTermTest(model04, ~ gender + edu)
regTermTest(model04, "gender")
regTermTest(model04, "edu")

```

# Taylor series linearization variance estimation method
## Frequency table, p. 19

```{r, warning=FALSE}
hints5_3_lin <- 
  as_survey_design(hints5_3, ids = VAR_CLUSTER, strata = VAR_STRATUM, 
                   weight = TG_all_FINWT0, nest = TRUE)
hints5_3_lin %>% 
  group_by(edu, gender) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE, vartype = c("se", "ci"))) %>% 
  drop_na() %>% 
  mutate_at(vars(starts_with("pct")), percent) %>% 
  kable() %>% 
  kable_styling()

```

## Chi-square tests, p. 20

```{r}
chisq_wald <- svychisq(~ edu + gender, hints5_3_lin, statistic = "Wald")
chisq_wald$statistic
chisq_wald$parameter

chisq_wald <- svychisq(~ edu + gender, hints5_3_lin, statistic = "adjWald")
chisq_wald$statistic
chisq_wald$parameter
```

## Merging HINTS survey iterations
### Assuming no differences between groups, pp. 52-54

```{r}
# Cycle 2, 100 replicate weights
# cycle2a, import Cycle 2 data
# cycle2b, new replicate weights 1 - 50: copy PERSON_FINWT1 - PERSON_FINWT50 as
#   Merged_NWGT1 - Merged_NWGT50
# cycle 2c, new replicate weights 51 - 100: copy PERSON_FINWT0 50 times as 
#   Merged_NWGT51 - Merged_NWGT100 
# temp_hints5_cycle2: join cycle2a, cycle2b, cycle2c by PersonID
cycle2a <- 
  read_sas(unz("HINTS_5_Cycle_2_SAS_03192020.zip",
               "HINTS 5- Cycle 2-SAS-03192020/hints5_cycle2_public.sas7bdat")) 
cycle2b <- cycle2a %>% 
  select(PersonID, c(paste0("PERSON_FINWT", 0:50))) %>% 
  mutate(Merged_NWGT0 = PERSON_FINWT0) %>% 
  select(-PERSON_FINWT0) %>% 
  rename_at(paste0("PERSON_FINWT", 1:50), ~ paste0("Merged_NWGT", 1:50))  
cycle2c <- map(1:50, ~ select(cycle2b, PersonID, Merged_NWGT0)) %>% 
  reduce(inner_join, by = "PersonID") %>% 
  set_names(c("PersonID", paste0("Merged_NWGT", 51:100)))
temp_hints5_cycle2 <- list(cycle2a, cycle2b, cycle2c) %>% 
  reduce(inner_join, by = "PersonID") %>% 
  mutate(survey = 1) 
# clean up
rm(cycle2a, cycle2b, cycle2c)

# Cycle 3, 100 replicate weights
# cycle3a, import Cycle 3 data
# cycle3b, new replicate weights 51 - 100: copy TG_all_FINWT1 - TG_all_FINWT50 as
#   Merged_NWGT51 - Merged_NWGT100
# cycle 3c, new replicate weights 1 - 50: copy TG_all_FINWT0 50 times as 
#   Merged_NWGT1 - Merged_NWGT50 
# temp_hints5_cycle3: join cycle3a, cycle3b, cycle3c by PersonID
cycle3a <- read_sas(unz("HINTS5_Cycle3_SAS_03112020.zip",
                         "hints5_cycle3_public.sas7bdat"))
cycle3b <- cycle3a %>% 
  select(PersonID, c(paste0("TG_all_FINWT", 0:50))) %>% 
  mutate(Merged_NWGT0 = TG_all_FINWT0) %>% 
  select(-TG_all_FINWT0) %>% 
  rename_at(paste0("TG_all_FINWT", 1:50), ~ paste0("Merged_NWGT", 51:100))
cycle3c <- map(1:50, ~ select(cycle3b, PersonID, Merged_NWGT0)) %>% 
  reduce(inner_join, by = "PersonID") %>% 
  set_names(c("PersonID", paste0("Merged_NWGT", 1:50)))
temp_hints5_cycle3  <- list(cycle3a, cycle3b, cycle3c) %>% 
  reduce(inner_join, by = "PersonID") %>% 
  mutate(survey = 2) 
rm(cycle3a, cycle3b, cycle3c)

# stack Cycle 2 and Cycle 3 into a single table and create survey object with
#   the new 100 replicate weights
# resulting table has 8,942 rows (3,504 for Cycle 2 and 5,438 for Cycle 3)
#   and 1,016 columns
#     914 variables appear in either or both of Cycle 2 and Cycle 3
#     102 added variables: cycle ID, final sample weight, 100 replicate
#     weights
mergeHINTSS5C2_HINTS5C3 <- 
  bind_rows(temp_hints5_cycle2, temp_hints5_cycle3) %>% 
  # reorder columns to put new weights in front
  select(survey, PersonID, num_range("Merged_NWGT", 0:100), everything()) %>% 
  # variable to distinguish survey iterations
  mutate_at("survey", factor, levels = 1:2,
            labels = c("HINTS 5 Cycle 2", "HINTS 5 Cycle 3")) %>% 
  # survey design object
  as_survey_rep(weights = "Merged_NWGT0",
                repweights = paste0("Merged_NWGT", 1:100), 
                type = "JK1", scale = 49/50, mse = TRUE)
rm(temp_hints5_cycle2, temp_hints5_cycle3)

```

### Frequencies on `SeekHealthInfo` and `ChanceAskQuestions`, p. 54

```{r}
mergeHINTSS5C2_HINTS5C3 %>% 
  mutate_at("SeekHealthInfo", factor, labels = c("NA", "Yes", "No")) %>% 
  group_by(SeekHealthInfo) %>% 
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE)) %>% 
  mutate_at(vars(starts_with("pct")), percent)
```


```{r}
mergeHINTSS5C2_HINTS5C3 %>%
  mutate_at("ChanceAskQuestions", factor,
            labels = c(paste0("NA", 1:5),
                       "Always", "Usually", "Sometimes", "Never")) %>%
  group_by(ChanceAskQuestions) %>%
  summarize(n = unweighted(n()),
            pct = survey_mean(na.rm = TRUE)) %>%
  mutate_at(vars(starts_with("pct")), percent)
```




